Presheaf Model for Dependent Types

### 2017-09-21 papers/talk-11/paper-11.tex
Rewrite introduction to dependent types

\usepackage{todonotes}
\usepackage{semantic}

@@

Dependent types are types that unlike simple types like functions and products
may depend on or vary with values.  A classical example of a dependent type is
a type of lists $List_{A}(N)$ of a given length $N : \NN$ over some
type $A$. The elements of such type are $Nil_{A} : List(0)$ and
$Cons_{A}(x, l) : List(N+1)$, where $x : {A}$ and $l : List_{A}(N)$.
Now let us consider a function that takes $n : \NN$ and returns a list of
length $n$ containing $0$ in all entries.  This function would have the type
$\Pi n : \NN.List_\NN(n)$, a type of functions for which the type
of the return value depends on the argument.  The important point of the
dependent typing is that it reveals more information about the function.
Another example is an exception-free head function that returns the first
element of the list:
\[
  hd: \Pi x : \mathbb{N}.List_{A}(Suc(x)) \to A
\]
The type of this function prevents it from being applied to an empty list,
showing clearly that the function is defined only for non empty lists.

@@

To formalize dependent types we build a system that allows us to make
judgements of the form $a : A$ (term $a$ has type $A$) and $A~type$ ($A$ is a
type). We will also want a notion of definitional equality to be built into the
system, for example we want to consider $0 : \NN$ and $0 + 0 : \NN$ to be
(definitionally) equal terms and $List_A(0)$, $List_A(0 + 0)$ as
(definitionally) equal types: if $l : List_A(0)$ then also $l : List_A(0 + 0)$.
This leads to two more forms of the judgements: $a = b : A$ ($a$ and $b$ are
definitionally equal terms of type $A$) and $A = B~type$ ($A$ and $B$ are
definitionally equal types). Finally, we must know types of the free variables
occurring in a judgement: we cannot assert $x + y : \NN$ unless we know $x :
\NN$ and $y : \NN$. Since such declarations may depend on each other like in $x
: \NN$, $y : List_\NN(x)$ it is convenient to make all judgements relative to a
list of variable declarations including at least the free variables occurring
inside the judgement. Such lists are called contexts and sometimes also type
assignments.

Intuitively, a context $x_1 : A_1, \dots, x_n : A_n$ is well-formed if each
$A_i$ is a type in the context $x_1 : A_1, \dots, x_{n-1} : A_{n-1}$ and all
the variables $x_i$ are pairwise distinct. Thus we arrive to six kinds of
judgements:

@@

We also have the following structural rules:
\begin{itemize}
  \item The empty context construction, context extension and context equality
    rules
    \[
      \inference[C-Emp]{}{\vdash \diamond~ctxt}
      \qquad
      \inference[C-Ext]{\Gamma \vdash A~type}{\vdash \Gamma, x : A~ctxt}
    \]
    \[
      \inference[C-Ext-Eq]{\vdash \Gamma = \Delta~ctxt \qquad \Gamma \vdash A = B~type}{\vdash \Gamma, x:A = \Delta, y:B~ctxt}
    \]
    where $x$ and $y$ are assumed to be fresh variables.
  \item The variable rule
    \[
      \inference[Var]{\vdash \Gamma, x:A,\Delta~ctxt}{\Gamma,x:A,\Delta \vdash x:A}
    \]
  \item Rules expressing that definitional equality is an equivalence relation:
    \[
      \inference[C-Eq-R]{\vdash \Gamma~ctxt}{\vdash \Gamma = \Gamma~ctxt}
    \]
    \[
      \inference[C-Eq-S]{\vdash \Gamma = \Delta~ctxt}{\vdash \Delta = \Gamma~ctxt}
    \]
    \[
      \inference[C-Eq-T]{\vdash \Gamma = \Delta~ctxt \qquad \vdash \Delta = \Theta~ctxt}{\vdash \Gamma = \Theta~ctxt}
    \]
    with analogous rules for terms and types (omitted).
  \item Rules relating typing and definitional equality:
    \[
      \inference[Tm-Conv]{\Gamma \vdash a : A \qquad \vdash \Gamma = \Delta~ctxt \qquad \Gamma \vdash A = B~type}{\Delta \vdash a : B}
    \]
    \[
      \inference[Ty-Conv]{\vdash \Gamma = \Delta~ctxt \qquad \Gamma \vdash A~type}{\Delta \vdash A~type}
    \]
  \item And rules for weakening and substitution, where $\mathcal{J}$ ranges over judgements of the form $a : A, A~type, a = b: A, A = B~type$:
    \[
      \inference[Weak]{\Gamma, \Delta \vdash \mathcal{J} \qquad \Gamma \vdash C~type}{\Gamma, x : C, \Delta \vdash \mathcal{J}}
    \]
    \[
      \inference[Subst]{\Gamma, x: C, \Delta \vdash \mathcal{J} \qquad \Gamma \vdash c: C}{\Gamma, \Delta[c/x] \vdash \mathcal{J}[c/x]}
    \]
    Here $\mathcal{J}[c/x]$ (and $\Delta[c/x]$) denotes the capture-free
    substitution of $x$ for $c$ in $\mathcal{J}$. This means that bound
    variables in $\mathcal{J}$ are systematically renamed to prevent any free
    variables in $c$ from becoming bound in $\mathcal{J}[c/x]$. The contexts,
    types, and terms will be considered equal if they agree up to names of
    bound variables and we will assume existence of capture-free substitution
    function on these equivalence classes.
\end{itemize}

\section*{Type formers}

Type and term formers are introduced by formation, introduction, elimination,
and equality rules. We present several type formers to give an example of how
they may be constructed.

@@

Dependent product (also known as dependent function space or $\Pi$-type)
corresponds to set-theoretic notion of cartesian product over a family of sets
$\Pi_{i \in I}B_i$ which has as elements functions mapping an index $i$ to an
element of a corresponding set $B_i$. In type theory this is expressed as
follows:

\[
  \inference[$\Pi$-F]{\Gamma \vdash A~type \qquad \Gamma, x:A \vdash B~type}{\Gamma \vdash \Pi x:A. B~type}
\]

\[
  \inference[$\Pi$-Eq]{\Gamma \vdash A_1 = A_2 ~type \qquad \Gamma, x:A \vdash B_1 = B_2 ~type}{\Gamma \vdash \Pi x:A_1. B_1 = \Pi x:A_2. B_2~type}
\]

The first rule expresses that a dependent product consists of a type $A$ (that
possibly depends on other types in $\Gamma$) and a type depending on $A$ (and
$\Gamma$), namely $B$. The rule $\Pi$-Eq expresses that the definitional
equality is respected by the $\Pi$-former. Note that the variable $x$ becomes
bound in $\Pi x:A. B$.

To form the elements of the $\Pi$-type we have the introduction rules with
associated congruence rule:

\[
  \inference[$\Pi$-I]{\Gamma, x:A \vdash b : B}{\Gamma \vdash \lambda x : A. b : \Pi x:A. B}
\]

\[
  \inference[$\Pi$-I-Eq]{\Gamma, x:A_1 \vdash b_1 = b_2 : B_1 \qquad \Gamma A_1 = A_2~type \qquad \Gamma x: A_1 \vdash B_1 = B_2~type}{\Gamma \vdash \lambda x : A_1. b^{B_1} = \lambda x : A_2. b^{B_2} : \Pi x:A_1. B_1}
\]

So to give an element of $\Pi x:A. B$ one must give an element of $B[x]$ in the
presence of a variable $x$ of type $A$. The congruence rule $\Pi$-I-Eq
expresses that definitional equality preserves $\Pi$-introduction. In the
following we will refrain from writing down the congruence rules explicitly,
having them present implicitly for every type and term formers we introduce.

Elements of $\Pi$-type are consumed using application:
\[
  \inference[$\Pi$-E]{\Gamma \vdash u : \Pi x:A. B \qquad \Gamma \vdash a : A}
  {\Gamma \vdash App_{[x:A]B}(u, a): B[a / x]}
\]

Square brackets in the typing annotation $[x : A] B$ here mean that the
variable $x$ is bound in $B$.

The following rule gives definitional equality to applications --- applying a
function $\lambda x : A. b^B$ to a term $a : A$ results in $b$ with $x$
replaced by $a$:

\[
  \inference[$\Pi$-C]{\Gamma \vdash \lambda x : A. b : \Pi x:A. B \qquad \Gamma \vdash a : A}
  {\Gamma \vdash App_{[x:A]B}(\lambda x : A. b, a) = b[a / x]: B [a / x]}
\]

Note that the substitution plays more prominent role here than in simply typed
lambda calculus: substitution is needed to formulate not only the term
conversion rules, but also the typing rules.

@@

The next type former we consider is the $\Sigma$-type, or dependent sum,
corresponding to disjoint union in set theory. If we are given a family of sets
$(B_i)_{i \in I}$ we can form the set $\Sigma_{i \in I} B_i := \{ (i, b)~|~i
\in I \land b \in B_i\}$ whose elements consist of an index $i$ and an element
of the corresponding set $B_i$. In type theory the corresponding formation and
introduction rules look as follows.
\[
  \inference[$\Sigma$-F]{\Gamma \vdash A \text{ type}\\ \Gamma, x:A \vdash B \text{ type}}{\Gamma \vdash \Sigma x : A. B \text{ type}}
\]
\[
  \inference[$\Sigma$-I]{\Gamma \vdash a : A\\ \Gamma\vdash b: B[a / x]}{\Gamma \vdash Pair_{[x:A]B} (a,b) : \Sigma x : A. B}
\]

Elimination rule:
\[
  \inference[$\Sigma$-E]{\Gamma, z : \Sigma x: A. B \vdash C \text{ type}\\ \Gamma, x: A, y: B \vdash c : C[Pair_{[x:A]B}(x, y)/z]\\ \Gamma \vdash u : \Sigma x: A. B}
  {R^\Sigma_{[z : \Sigma x: A. B]C}([x : A, y: B]c, u): C[u/z]}
\]

And definitional equality:
\[
  \inference{\Gamma\vdash R^\Sigma_{[z : \Sigma x: A. B]C}([x : A, y: B]c, Pair_{[x:A]B}(a,b)): C[Pair_{[x:A]B}/z]}
  {\Gamma\vdash R^\Sigma_{[z : \Sigma x: A. B]C}([x : A, y: B]c, Pair_{[x:A]B}(a,b)) = c[a/x,b/y] : C [Pair_{[x:A]B}/z}
\]

The projection functions for the dependent sum can be defined using these
rules:
\[
  \mathbf{p}(a) := R^\Sigma_{[z : \Sigma x: A. B]A}([x : A, y: B]x, a) : A
\]
for the first projection and
\[
  \mathbf{q}(a) := R^\Sigma_{[z : \Sigma x: A. B]B[z.1/x]}([x : A, y: B]y, a) : B[\mathbf{p}(a)/x]
\]
for the second projection, assuming $\Gamma \vdash A~type$, $\Gamma, x:A \vdash
B~type$, and $\Gamma \vdash a : \Sigma x: A. B$.

The important special cases of $\Pi x: A. B$ and $\Sigma x : A. B$ arise when
$B$ does not actually depend on $A$. The we write
\[
  A \to B := \Pi x: A. B
\]
and
\[
  A \times B := \Sigma x: A. B
\]
indicating that in this cases $\Pi$- and $\Sigma$-types correspond to ordinary
non-dependent function and cartesian product types respectively.

@@

We will use categories with families (CwF) to model dependent types. First
consider what is a category of families. The category of families of sets,
$\fcat{Fam}$, has as objects $(A, B)$ where $A$ is a set and $B = (B_a~|~a \in
A)$ is an $A$-indexed family of sets $B_a$, $a \in A$. A morphism $(A, B) \to
(A', B')$ is given by a pair $(f, g)$ where $f: A \to A'$ and $g$ is an
$A$-indexed famiy of functions such that $g_a: B_a \to B_{fa}'$.

A \emph{category with families} is given by ($\cat{C}$,$F$) where:

@@

  \item[$\bullet$] There is also the operation of \emph{context extension}: if
    $\vdash \Gamma~ctxt$ and $\Gamma \vdash A~type$, there is a context
    $\Gamma, x:A$ (for brevity also written as $\Gamma.A$ from now on), a
    context morphism $\mathbf{p}: \Gamma.A \to \Gamma$, and a term $\Gamma.A
    \vdash \mathbf{q} : A \mathbf{p}$. This should satisfy the following
    property: for each $\vdash \Delta~ctxt$ and a context morphism
    $\sigma \from \Delta \to \Gamma$ and $\Delta \vdash a : A \sigma$, there is
    a context morphism $(\sigma, a): \Delta \to \Gamma.A$ such that:
    \[
      \mathbf{p}(\sigma, a) = \sigma \quad \mathbf{q}(\sigma, a) = a \quad (\sigma, a) \tau =
      (\sigma \tau, a \tau) \quad (\mathbf{p}, \mathbf{q}) = 1
    \]
    Where in the last equation $1: \Gamma.A \to \Gamma.A$
\end{itemize}

We say that a category with families supports $\Pi$- and $\Sigma$-types if it's
closed under the corresponding rules shown in previous sections.

>>>>

\usepackage{todonotes}
\usepackage{semantic}
\usepackage{mathpartir}

@@

Suppose you want to write a function that returns the head of a given list of
natural numbers, using the classical definition of the list type:
\[
  List = Nil : Unit \to List~|~Cons : \NN \times List \to List
\]
Writing the case for $Cons$ is easy --- just take the first element of the
pair.  But what about the case of an empty list? This case is tricky because it
is not clear what should the result be. Obviously, such function (if we do not
allow definition of partial functions which is the case in many programming
languages) cannot be well defined on the empty list without taking arbitrary
decisions. But maybe such problems can be prevent from happening at all?

A solution to this issue is an instance for the application of dependent types.
Dependent types are such types that include simple types, like products and
functions, that we have seen in simply-typed lambda calculus (STLC), but also
add a possibility to define types using values. With dependent types we would
be able to define a type for lists that includes the size of the list:
\[
  List(\NN) = Nil : Unit \to List(0)~|~Cons : \NN \times List(n) \to List(Suc(n))
\]
\begin{example}
  The function to construct a list of length $n$, having only $0$ as elements
  would then have the type $\Pi n: \NN.List(n)$, indicating that it accepts a
  natural number and returns a value of a type constructed using that number.
\end{example}
\begin{example}
  Our desired head function would have the type
  \[
    Hd: \Pi n: \NN.List(Suc(n)) \to \NN
  \]
  indicating that it cannot be applied to the empty list at all.
\end{example}

The important part is that the dependent typing reveals more information about
the function.

@@

To formalize dependent types we will use a notion of typing context (which is
almost the same as typing environment as seen for STLC, with the exception that
it holds dependent types).

\begin{definition}[Context]
  A \emph{context} is a list of bindings that connect variables to types: $x_1: A_1,
  \dots, x_n: A_n$. A context is called well-formed if every variable in it is
  distinct from the other variables and each $A_i$ is a type in the context $x_1
  : A_1, \dots, x_{n-1} : A_{n-1}$, since our types may now include the variables
  from the context.
\end{definition}

Note that we extend the notion of a well-formed context from STLC --- in STLC
we only needed the variables to be distinct, since they had no influence on the
types.

Having contexts, we want to work with the following judgements: $a : A$ (term
$a$ has type $A$) and $A~type$ ($A$ is a type) and consider definitional
equality to be built-in into our formal system. That is, we want the types
$List(0)$ and $List(0 + 0)$ to be equal. So we also include the judgements
about the equality of terms $a = b : A$, types $A = B~type$, and contexts
$\Gamma = \Delta~ctxt$.

To put it together, we have six types of judgements:

@@

We also need the structural rules:

\begin{definition}[Structural rules]
\begin{mathpar}
  \inferrule*[right=C-Emp]{ }{\vdash \diamond~ctxt}
  \and
  \inferrule*[right=C-Ext]{\Gamma \vdash A~type}{\vdash \Gamma, x : A~ctxt}
  \and
  \inferrule*[right=C-Ext-Eq]{\vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A = B~type}{\vdash \Gamma, x:A = \Delta, y:B~ctxt}
  \and
  \inferrule*[right=Var]{\vdash \Gamma, x:A,\Delta~ctxt}{\Gamma,x:A,\Delta \vdash x:A}
  \and
  \inferrule*[right=C-Eq-R]{\vdash \Gamma~ctxt}{\vdash \Gamma = \Gamma~ctxt}
  \and
  \inferrule*[right=C-Eq-S]{\vdash \Gamma = \Delta~ctxt}{\vdash \Delta = \Gamma~ctxt}
  \and
  \inferrule*[right=C-Eq-T]{\vdash \Gamma = \Delta~ctxt \\ \vdash \Delta = \Theta~ctxt}{\vdash \Gamma = \Theta~ctxt}
  \and
  \inferrule*[right=Tm-Conv]{\Gamma \vdash a : A \\ \vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A = B~type}{\Delta \vdash a : B}
  \and
  \inferrule*[right=Ty-Conv]{\vdash \Gamma = \Delta~ctxt \\ \Gamma \vdash A~type}{\Delta \vdash A~type}
  \and
  \inferrule*[right=Weak]{\Gamma, \Delta \vdash \mathcal{J} \\ \Gamma \vdash C~type}{\Gamma, x : C, \Delta \vdash \mathcal{J}}
  \and
  \inferrule*[right=Subst]{\Gamma, x: C, \Delta \vdash \mathcal{J} \qquad \Gamma \vdash c: C}{\Gamma, \Delta[c/x] \vdash \mathcal{J}[c/x]}
\end{mathpar}

where $x$ and $y$ in C-Ext and C-Ext-Eq are assumed to be fresh variables and
$\mathcal{J}$ in Weak and Subst ranges over judgements of the form $a : A,
A~type, a = b: A, A = B~type$. $[c/x]$ denotes the same substitution as seen in
STLC, except that it now also applies to the variables that may be contained in
types.
\end{definition}

The rules C-Emp and C-Ext give us the empty context ($\diamond$) and allow us
to extend the context (analogous to the environments from STLC).

C-Ext-Eq gives the meaning to the context equality: if we have two equal
contexts and one of them contains equal types, then the variables of these
types are the same in both contexts.

Var says that if a context contains a variable of a certain type, then the
judgement that this variable has that type in the context is satisfied.

Rules C-Eq define that context equality is an equivalence relation, very
similar rules for terms and types, stating that their equalities are also
equivalence relations, are omitted for brevity.

The rule Tm-Conv states that equivalent types and equivalent contexts may be
used in place of each other in judgements on terms, while the rule Ty-Conv
states that equivalent contexts can be used in place of each other in
judgements on types.

The rule Weak says that we can extend our context with a valid type, preserving
the validity of the typing judgements.

The rule Subst states that the validity of the typing judgements is preserved
under substitution.

There are two important dependent type formers to consider: dependent product
and dependent sum. The types formed with them have well-understood analogues in
set theory and are essential when working with dependent types. These are not
the only type formers that we can construct and we hope that the general
principles of their construction will be clear from these two examples.

To present these type formers we will be using formation, introduction,
elimination, and equality rules.

@@

Dependent product or dependent function type (often abbreviated as $\Pi$-type)
is a type of a function that returns a result which type may depend on the
function argument. It is defined by the following rules:

\begin{definition}[Dependent product]
  \begin{mathpar}
    \inferrule*[right=$\Pi$-F]{\Gamma \vdash A~type \\ \Gamma, x:A \vdash B~type}{\Gamma \vdash \Pi x:A. B~type}
    \and
    \inferrule*[right=$\Pi$-Eq]{\Gamma \vdash A_1 = A_2 ~type \\ \Gamma, x:A_1 \vdash B_1 = B_2 ~type}{\Gamma \vdash \Pi x:A_1. B_1 = \Pi x:A_2. B_2~type}
    \and
    \inferrule*[right=$\Pi$-I]{\Gamma, x:A \vdash b : B}{\Gamma \vdash \lambda x : A. b : \Pi x:A. B}
    \and
    \inferrule*[right=$\Pi$-I-Eq]
      {\Gamma, x:A_1 \vdash b_1 = b_2 : B_1 \\ \Gamma \vdash A_1 = A_2~type \\ \Gamma, x: A_1 \vdash B_1 = B_2~type}
      {\Gamma \vdash \lambda x : A_1. b^{B_1} = \lambda x : A_2. b^{B_2} : \Pi x:A_1. B_1}
    \and
    \inferrule*[right=$\Pi$-E]{\Gamma \vdash u : \Pi x:A. B \\ \Gamma \vdash a : A}{\Gamma \vdash App_{[x:A]B}(u, a): B[a / x]}
    \and
    \inferrule*[right=$\Pi$-C]{\Gamma \vdash \lambda x : A. b : \Pi x:A. B \\ \Gamma \vdash a : A}
    {\Gamma \vdash App_{[x:A]B}(\lambda x : A. b, a) = b[a / x]: B [a / x]}
  \end{mathpar}
\end{definition}

The rule $\Pi$-F serves for the formation of the type: if we know that having a
variable $x: A$ we obtain a type $B$, then we have our $\Pi$-type, making $x$
bound in that type.

The rule $\Pi$-Eq expresses that the definitional equality is respected by the
$\Pi$-former.

To introduce instances of $\Pi$-type we use the rule $\Pi$-I which looks almost
the same as in the simply-typed case, with the exception that our $\lambda$
here forms a dependent function type, not the simply-typed one.

The rule $\Pi$-I-Eq serves similar purpose to $\Pi$-Eq and ensures the
preservation of the definitional equality. The similar equality preservation
rules henceforth will be omitted for the other rules we introduce, but the
reader should keep their implicit presence in mind while seeing other term and
type formers that we introduce.

The rule $\Pi$-E regulates the application of a dependent function to some value
of the correct type results in the return value of a proper type. Note the rule
only concerns the resulting type here, not the value and that the bound
variable $x$ (shown as $[x : A] B$ in subscript) is replaced in the type, not
the value, which is handled by the next rule.

The rule $\Pi$-C defines the resulting value of a dependent function
application. It is almost the same $\beta$-conversion rule as in STLC, with the
variable substitution also occurring in the type.

The analogue to dependent product in set theory is a cartesian product over a
family of sets $\Pi_{i \in I}B_i$ where elements are the functions mapping an
index $i$ to an element of a corresponding set $B_i$.

@@

The other very important type former is dependent sum type (also known as
dependent pair or $\Sigma$-type).

\begin{definition}[Dependent sum]
  \begin{mathpar}
    \inferrule*[right=$\Sigma$-F]{\Gamma \vdash A \text{ type}\\ \Gamma, x:A \vdash B \text{ type}}{\Gamma \vdash \Sigma x : A. B \text{ type}}
    \and
    \inferrule*[right=$\Sigma$-I]{\Gamma \vdash a : A\\ \Gamma\vdash b: B[a / x]}{\Gamma \vdash Pair_{[x:A]B} (a,b) : \Sigma x : A. B}
    \and
    \inferrule*[right=$\Sigma$-E]
      {\Gamma, z : \Sigma x: A. B \vdash C \text{ type} \\ \Gamma, x: A, y: B \vdash c : C[Pair_{[x:A]B}(x, y)/z] \\ \Gamma \vdash u : \Sigma x: A. B}
      {R^\Sigma_{[z : \Sigma x: A. B]C}([x : A, y: B]c, u): C[u/z]}
    \and
    \inferrule*[right=$\Sigma$-C]
      {\Gamma\vdash R^\Sigma_{[z : \Sigma x: A. B]C}([x : A, y: B]c, Pair_{[x:A]B}(a,b)): C[Pair_{[x:A]B}/z]}
      {\Gamma\vdash R^\Sigma_{[z : \Sigma x: A. B]C}([x : A, y: B]c, Pair_{[x:A]B}(a,b)) = c[a/x,b/y] : C [Pair_{[x:A]B}/z]}
  \end{mathpar}
\end{definition}

The formation rule $\Sigma$-F follows the same pattern as the rule for
dependent function, but the introduction is a bit different. Note that we take
such $b$, where the type contains a variable $x : A$ that is substituted by the
value $a$. Together the values $a$ and $b$ form a pair, where the type of the
second element depends on the value of the first. We use the notation $[x : A]
B$ here again to indicate that the variable $x$ is bound in the type $B$.

The rule $\Sigma$-E is again analogous to the one seen in dependent product: we
can replace a variable of a dependent pair type with a concrete term if he
have a term that contains that variable. And again we just say what happens to
the type, not the value there.

To define the effect on the values we use the rule $\Sigma$-C, which says that
the elements of the pair will replace the corresponding variables in the body
of the given term, but not the pair variable itself, such substitution occurs
only in the type.

We know the projection functions for the simply-typed products in STLC. Let us
now define the projection functions for the \emph{dependent} products, which we
call $\mathbf{p}$ for the first projection and $\mathbf{q}$ for the second
projection:
\[
  \mathbf{p}(a) := R^\Sigma_{[z : \Sigma x: A. B]A}([x : A, y: B]x, a) : A
\]
\[
  \mathbf{q}(a) := R^\Sigma_{[z : \Sigma x: A. B]B[z.1/x]}([x : A, y: B]y, a) : B[\mathbf{p}(a)/x]
\]
where $\Gamma \vdash A~type$, $\Gamma, x:A \vdash B~type$, and $\Gamma \vdash a
: \Sigma x: A. B$.

The analogue to dependent sum in set theory is a disjoint union. For every
family of sets $(B_i)_{i \in I}$ we can form the set $\Sigma_{i \in I} B_i :=
\{ (i, b)~|~i \in I \land b \in B_i\}$ whose elements consist of an index $i$
and an element of the corresponding set $B_i$.

It is worth notice that we may define the simply-typed products and functions
in terms of dependent types: we just need the types to not depend on each
other. So, assuming that $B$ does not depend on $A$, we get:
\[
  A \to B := \Pi x: A. B
\]
and
\[
  A \times B := \Sigma x: A. B
\]

@@

There are several ways to model dependent types with category theory, the one
we show is using Categories with Families.

\begin{definition}[Category of families of sets]
  The category of families of sets, $\fcat{Fam}$, has as objects $(A, B)$ where
  $A$ is a set and $B = (B_a~|~a \in A)$ is an $A$-indexed family of sets $B_a$,
  $a \in A$. A morphism $(A, B) \to (A', B')$ is given by a pair $(f, g)$ where
  $f: A \to A'$ and $g$ is an $A$-indexed famiy of functions such that $g_a: B_a
  \to B_{fa}'$.
\end{definition}

\begin{definition}[Category with Families (CwF)]
A \emph{category with families} is given by ($\cat{C}$,$F$) where:

@@

  \item[$\bullet$] There is also the operation of \emph{context extension}: if
    $\vdash \Gamma~ctxt$ and $\Gamma \vdash A~type$, there is a context
    $\Gamma, x:A$ (for brevity also written as $\Gamma.A$ from now on), a
    context morphism $\mathbf{p}: \Gamma.A \to \Gamma$, and a term $\Gamma.A
    \vdash \mathbf{q} : A \mathbf{p}$. This should satisfy the following
    property: for each $\vdash \Delta~ctxt$ and a context morphism
    $\sigma \from \Delta \to \Gamma$ and $\Delta \vdash a : A \sigma$, there is
    a context morphism $(\sigma, a): \Delta \to \Gamma.A$ such that:
    \[
      \mathbf{p}(\sigma, a) = \sigma \quad \mathbf{q}(\sigma, a) = a \quad (\sigma, a) \tau =
      (\sigma \tau, a \tau) \quad (\mathbf{p}, \mathbf{q}) = 1
    \]
    Where in the last equation $1: \Gamma.A \to \Gamma.A$
\end{itemize}
\end{definition}

We say that a category with families supports $\Pi$- and $\Sigma$-types if it is
closed under the corresponding rules shown in previous sections.
